<!--
Analysis of anthrax infected cells vs. control cells
-->
```{r setup, echo=FALSE}
opts_chunk$set(tidy=TRUE, highlight=TRUE, fig.align='left', fig.show='hold',
               cache=FALSE, highlight=TRUE, autodep=TRUE, warning=FALSE, error=FALSE,
               message=FALSE, prompt=TRUE, comment='', fig.cap='')
```

# Overview
Infecting mouse lung macrophages with influenza causes a decrease in
the antibacterial function of the lungs which bottoms out at D9 but
improves closer to D11. The group is interested in seeing which sets
of genes might be involved in that process.

The data is two sets of six replicates of RNA-seq data, one set of six
samples at D9 and the other six at D11. These samples were barcoded
and pooled and run in lanes together. Overall two identical lanes were
run and the reads were combined.

There are two 'experiments' in this dataset which gives it a little
messier structure than it would have normally: in one experiment, only
the RNA was extracted and one which had the miRNA and RNA
extracted. Also for each of the two experiments the RNA was extracted
on a different day, so there is a little bit of a confound between day
of extraction and type of extraction. I codified this information in
the data/macrophage.csv file so we can take it into account during the
analysis. It is not clear if these were also from the same litter of
mice as well; at any rate we should expect to see some variation
across the two groups.

The RNA was extracted from FACS sorted macrophages, I don't know if that
means they need to do some kind of amplification of the RNA as well or if
they got enough RNA just from the sort.

They also have qPCR data regarding the regulation of MARCO, a phagocytic
receptor on the macrophages and have shown there is a large increase
in D9-D11, so we should see that too.

```{r load-data}
library(CHBUtils)
library(edgeR)
library(HTSFilter)
library(ggplot2)
library(gridExtra)
wd = "/Users/rory/Volumes/odyssh/hsph/projects/kobzik_macrophage"
setwd(wd)
metadata_file = "data/macrophage.csv"
metadata = read.csv(metadata_file, header=TRUE, colClasses="factor")

count_file = "macrophage/final2/131126_macrophage/combined.counts"
counts = read.table(count_file, header=TRUE, sep="\t")
colnames(counts) = gsub(".", "_", colnames(counts), fixed=T)
rownames(counts) = counts$id
counts$id = NULL
samples = data.frame(samplename=colnames(counts))
samples = merge(samples, metadata, by="samplename", sort=FALSE)
MARCO_ID = "ENSMUSG00000026390"
```

# Overview
The MDS plot is kind of all over the place both looking at the day and the extraction,
but since this isn't cell-line data that kind of variability is not uncommon.
```{r mds-plot}
p1 = mds(counts, samples$day)
p2 = mds(counts, samples$extraction)
p3 = variance_by_component(counts)
grid.arrange(p1, p2, p3)
```

# Design
We'll use a blocking structure which will take into account the different extraction
days and protocols.
```{r design}
rownames(samples) = samples$samplename
design = model.matrix(~ 0 + day, data=samples)
#design = model.matrix(~0 + day + extraction, data=samples)
design
```

# Dispersion estimation
It is useful to look at the variation in the data and make sure it is
within the bounds we expect and to put some realistic expectation on
what kind of calls we can make. You can see from the dispersion plot
that genes with low CPM (counts per million) have a high degree of
variation-- a small signal with a large variation means that we won't
be able to make reliable differential calls on most of these genes. A
common thing people do is to just remove those genes from
consideration entirely, but I don't like doing that since it is
possible there is a reliable signal or two lurking there.  As long as
we keep in mind that those low CPM genes are off in the weeds a bit,
it is fine.

```{r dispersions}
y = DGEList(counts = counts)
y = estimateGLMCommonDisp(y, design)
y = estimateGLMTrendedDisp(y, design)
y = estimateGLMTagwiseDisp(y, design)
bcv(y)
```

One thing we will do later is to try to cut down on the number of
genes we are considering for differntial expression. In a single cell,
somewhere around 1/3 to a half of all possible genes are expressed, so
that means that there are a low number of genes which all will be
considered for differential expression but aren't changing because
they aren't even expressed. We can remove these low count genes via a
couple of different methods.  The one most people do is to just set a
threshold of 2-3 log CPM and only consider those genes, which I don't
like because you throw out some signal.  An alternative method would
be to keep only genes that could possibly be DE, that is genes which
are not lowly expressed with a similar expression level across all
conditions. There is a package to do that which we'll use. You can see
the effect of what it does by looking at the BCV plot after they are
removed. A lot of the lowly expressed genes have been removed as well as
some of the lowly expressed extremely variable genes.

```{r htsfilter-preview}
bcv(HTSFilter(y)$filteredData)
```

# Differential expression
```{r differential-expression}
fit = glmFit(y, design)
day9_vs_day11 = makeContrasts(day9_vs_day11=day11-day9, levels=design)
lrt = glmLRT(fit, contrast=day9_vs_day11)
topTags(lrt)
lrt_filt = HTSFilter(lrt, DGEGLM=fit, s.len=25, plot=FALSE)$filteredData
topTags(lrt_filt)
```

Unfortunately we aren't picking up anything from this analysis. Since there
didn't seem to be any systematic differences between the experiment 1 and
experiment 2 samples, we could try looking at that simpler model.

```{r differential-expression-two-factor}
design = model.matrix(~0 + day, data=samples)
design
y = DGEList(counts = counts)
y = estimateGLMCommonDisp(y, design)
y = estimateGLMTrendedDisp(y, design)
y = estimateGLMTagwiseDisp(y, design)
bcv(y)
fit = glmFit(y, design)
day9_vs_day11 = makeContrasts(day9_vs_day11=day11-day9, levels=design)
lrt = glmLRT(fit, contrast=day9_vs_day11)
topTags(lrt)
lrt_filt = HTSFilter(lrt, DGEGLM=fit, s.len=25, plot=FALSE)$filteredData
topTags(lrt_filt)
```

Still nothing is popping out as significant. EdgeR and other count-based methods are not
the best at identifying DE genes that are expressed in only one condition, so maybe
voom + limma will be better.

```{r voom}
library(limma)
#design = model.matrix(~0 + day + extraction, data=samples)
design = model.matrix(~day, data=samples)
nf = calcNormFactors(counts, method = "TMM")
y = calcNormFactors(y)
v = voom(y, design, plot=TRUE)
fit = lmFit(v, design)
fit = eBayes(fit)
voom.pvalues = fit$p.value[, 2]
voom.adjpvalues = p.adjust(voom.pvalues, method = "BH")


design = model.matrix(~day, data=samples)
voom.data = voom(counts, design = design, lib.size = colSums(counts) * nf)
voom.data$genes = rownames(counts)
voom.fitlimma = lmFit(voom.data, design = design)
voom.fitbayes = eBayes(voom.fitlimma)
voom.pvalues = voom.fitbayes$p.value[, 2]
voom.adjpvalues = p.adjust(voom.pvalues, method = "BH")
head(sort(voom.adjpvalues))
```

It is still not looking good, those are the lowest adjusted p-values from
voom + limma.

# Why aren't we seeing anything?

The correlations between the samples are lower than we would
expect. You can see that sample D11_3 looks like an outlier, though
the scale on the heatmap is a little bit misleading, however the
correlations between most of the samples are not very good in general.
That is reflected in the coefficient of biological variation plot
above, the common dispersion line (the horizontal line) is higher than normal,
it should be closer to 0.4 or 0.5 for these types of samples. A heatmap
of the pairwise correlations of the normalized counts of each sample is below.

```{r correlations}
library(gplots)
nc = cpm(y)
cor(nc)
heatmap.2(cor(nc), trace="none")
```

If we cluster the samples by MARCO fold change, there are two of the
six samples from each day which don't cluster with the of the samples
from their day.

```{r marco}
marco_expr = nc[MARCO_ID,]
marco_expr
heatmap.2(log(outer(marco_expr, marco_expr, "/")), trace="none")
```

Does looking at ACTB fold change show a similar clustering effect? 
```{r actb}
actb_expr = nc["ENSMUSG00000029580",]
heatmap.2(log(outer(actb_expr, expr_expr, "/")), trace="none")
```

No, and the correlations are much higher, this looks more randomly
distributed. 

The MARCO clustering is similar to the result found when the data was
analyzed before using FPKM from Cufflinks. The samples that don't
cluster with MARCO fold change (D9_1, D9_6, D11_1, D11_2) have some
overlap with the set
of samples which don't cluster properly using all of the
genes (D9_6, D11_1, D11_2), but not quite completely overlaping. I don't really like
removing samples that don't cluster together and then calling
differential expression on them, because it is cheating unless you
have a good reason for it. You are basically saying, I think there are
systematic differences based on a factor between these samples, let's
first remove the samples that would make that hypothesis invalid, and
then test for differential expression. That isn't proper.

In this case having the MARCO marker gives us an independent reason for
dropping samples if we choose to go that route.

For now let's try using the MARCO expression in the RNA-seq data as a
factor to consider. We'll rescale the MARCO expression vector so it is
centered on zero and are in units of SD from zero and add that into the
experimental design:

```{r marco_center}
marco_rescale = (marco_expr - mean(marco_expr)) / sd(marco_expr)
samples$standard_marco = marco_rescale
samples
```

And run differential expressing using our rescaled MARCO values as a regressor.
The way to interpret the log fold change of this are these are the fold change in
a gene that occurs for each standard deviations MARCO deviates from the mean.

```{r marco-analysis}
keep = rowSums(cpm(counts)) > 3
design = model.matrix(~day + marco_rescale, data=samples)
z = y[keep,]
z = estimateGLMCommonDisp(z, design)
z = estimateGLMTrendedDisp(z, design)
z = estimateGLMTagwiseDisp(z, design)
fit = glmFit(z, design)
lrt = glmLRT(fit, coef=3)
topTags(lrt)

ensembl_gene = "mmusculus_gene_ensembl"
filter_type = "ensembl_gene_id"
gene_symbol = "mgi_symbol"
annotate_df = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, gene_symbol, "description"),
		filters=c(filter_type), values=d[,"id"],
		mart=ensembl)
	m = merge(d, a, by.x='id', by.y=filter_type)
	return(m)
}
tags = topTags(lrt, dim(lrt$fitted.values)[1])$table
tags$id = rownames(tags)
tags = annotate_df(tags)
tags = tags[order(tags$PValue),]
```

```{r table-output, results='asis'}
library(googleVis)
print(gvisTable(head(tags, 100)),  options=list(width=800, height=500)), "chart")
write.table(tags, file="FC_vs_unit_of_marco_change.tsv", quote=FALSE,
  sep="\t", row.names=FALSE)
```

This spits out a promising list of genes, although none are significantly
different statistically. The table above is supposed to scroll around, but
something isn't working right with the way Google has been doing it. You might
have to click in the box and scroll around or possibly use the arrow keys to
scroll around, I haven't figured out what the bug is. I included the top 100 in
the table and saved the entire list of genes to the file 'FC_vs_unit_of_marco_change.tsv'.


# D9 vs D11 with dropping samples
For this, lets try dropping the four samples which don't cluster with
the others in their group looking at MARCO fold change derived from
the sequencing data (D9_1, D9_6, D11_1, D11_2). That way our group is
D9 vs D11, looking at only the low MARCO expressors in D9 and the high
MARCO expressors in D11. Since we already rigged it to be high and low
MARCO expressors, we'll just use day to run the model.

```{r D9_vs_D11_low_vs_high_only}
keep_samples = !colnames(z) %in% c("D9_1", "D9_6", "D11_1", "D11_2")
z = z[,keep_samples]
z = estimateGLMCommonDisp(z, design)
z = estimateGLMTrendedDisp(z, design)
z = estimateGLMTagwiseDisp(z, design)
design = model.matrix(~day, data=samples[keep_samples,])
fit = glmFit(z, design)
lrt = glmLRT(fit)
topTags(lrt)

ensembl_gene = "mmusculus_gene_ensembl"
filter_type = "ensembl_gene_id"
gene_symbol = "mgi_symbol"
annotate_df = function(d) {
	require(biomaRt)
	ensembl = useMart('ensembl', dataset = ensembl_gene)
	a = getBM(attributes=c(filter_type, gene_symbol, "description"),
		filters=c(filter_type), values=d[,"id"],
		mart=ensembl)
	m = merge(d, a, by.x='id', by.y=filter_type)
	return(m)
}
tags = topTags(lrt, dim(lrt$fitted.values)[1])$table
tags$id = rownames(tags)
tags = annotate_df(tags)
tags = tags[order(tags$PValue),]
```


cluster with MARCO fold change (D9_1, D9_6, D11_1, D11_2)is almost the
# TODO
I know this isn't exactly what you were looking for, and I'll do the D9-D11 difference
too, but  I want to see if we're on the right track before thinking about how to do it.
Is the key question you want to know what is different between D9 or D11, or is it
which genes are correlated with high MARCO expression? Not all of the D9 and D11 samples
have low or high MARCO expression, so to do the day plus MARCO expression, we'll have to
drop some of the replicates.
